% --- 摘要页设置开始（不另起页，避免首页空白） ---
% 垂直间距微调（缩小则中英文摘要块更紧凑；单位可用 cm / em）
\newcommand{\AbstractTopVSpace}{0.55cm}      % 页内「摘要/ABSTRACT」标题上方留白（原约 1cm）
\newcommand{\AbstractTitleBodyVSpace}{0.45em}% 标题与正文之间（原 0.8em）
\newcommand{\AbstractBodyKeywordVSpace}{0.45em}% 正文末与「关键词」之间（原 0.8em）
\pagenumbering{Roman} % 设置页码为罗马数字 I, II, III
\setcounter{page}{1} % 重置页码为 I
\pagestyle{front}    % 摘要、目录页脚：小五号 Times New Roman 罗马数字
\markboth{摘要}{摘要}
% 先在目录中登记“摘要”，确保目录中的页码为 I（中文摘要首页），
% 且作为与“第一章 绪论”同级的目录项（使用 section 级别以统一缩进和间距）
\phantomsection
\addcontentsline{toc}{section}{摘要}
% 抑制首页可能出现的脚注标记等多余输出
\makeatletter
\let\@makefnmark\relax
\renewcommand\@makefntext[1]{\relax}
\makeatother

% --- 中文摘要：标题“摘要”与章标题一致（三号黑体居中），正文与正文部分一致（小四宋体、首行缩进2字符、1.5倍行距）---
\vspace{\AbstractTopVSpace}
\begin{center}
    \centering
    {\sanhao\heiti 摘\quad 要}
\end{center}
\vspace{\AbstractTitleBodyVSpace}

% 正文首行缩进
\setlength{\parindent}{2em}
{\fontsize{12pt}{18pt}\selectfont\songti
推荐系统已从单纯的流量分发转向对用户深度兴趣的挖掘。异构信息网络（Heterogeneous Information Network）能够整合多种异构数据源、挖掘隐藏的语义信息，是提升推荐精度的可行途径。同时，在《个人信息保护法》与 GDPR 的合规约束下，加之各平台间“数据围墙”加高，传统的集中式大图训练模式难以同时满足跨域融合与隐私合规。联邦学习（Federated Learning, FL）以“数据不动模型动”的方式协同训练，参与方在本地保留原始数据、仅交换经保护的模型参数，有利于满足合规要求。然而在联邦场景下，现有异构图推荐研究仍面临冷启动、推荐性能受限、通信瓶颈与工程落地难等问题。

围绕上述问题，本文从推荐效果、通信效率与工程落地三方面展开。针对冷启动与 Non-IID 导致的效果退化，提出联邦异构图推荐框架 FedASCL，在客户端以属性与结构双视图对比学习补偿零交互用户的语义缺失，在服务器以全局语义原型对齐抑制客户端漂移。针对端云通信瓶颈，进一步提出语义感知的模型参数压缩策略，在约 $15.2\times$ 通信缩减下以可接受的精度回落换取带宽收益。最后设计并实现面向学术场景的联邦论文推荐原型系统，将 FedASCL 与压缩策略集成进端云协同流程。算法层面在 MovieLens-1M、Yelp 与 ACM 等公开基准上完成离线验证，系统层面通过功能测试与通信仿真完成工程可行性验证。主要工作包括：
\begin{cnenumerate}
    \cnitem 针对冷启动与数据异质性，提出基于属性-结构双视图对比学习的联邦异构图推荐框架 FedASCL（Federated Attribute-Structure Contrastive Learning）：在潜在空间中通过跨视图对比学习对齐用户属性与图结构语义，以属性特征补偿交互缺失；并引入全局语义原型对齐（Global Semantic Prototype Alignment）约束本地表征、缓解客户端模型漂移（Client Drift）。在第三章冷启动设定下相对次优基线提升约 3\%--8\%，在极端 Non-IID（$\alpha=0.1$）设定下 NDCG@10 相对次优基线约提升 8.2\%、相对 IID 的衰减（5.4\%）不足次优基线（13.6\%）的一半。
    \cnitem 为降低通信开销，提出语义感知的模型参数压缩策略，设计轻量级元路径选择器在本地训练中动态剔除低贡献冗余通道，并采用残差梯度量化压缩传输。第四章在更严联邦设定（$N{=}100$、$\alpha{=}0.5$）下，于 MovieLens-1M、Yelp 与 ACM 三数据集上开展实验。约 $15.2\times$ 通信缩减时，Recall@20 相对 Full 在 MovieLens-1M、Yelp 与 ACM 上分别回落约 2.4\%、3.5\% 与 3.2\%，累计通信流量降低约一个数量级。
    \cnitem 针对联邦推荐系统落地难的问题，本文设计并实现了面向学术推荐场景的联邦论文推荐原型系统。系统采用端云协同的分层架构，贯通模型分发、本地 FedASCL 训练、语义感知压缩上传与安全聚合等关键环节；业务层支持异步任务调度，联邦训练轮次采用同步聚合，与第三、四章实验设定保持一致。系统提供可配置的本地差分隐私模块，并通过功能测试与通信仿真验证了模块集成与带宽优化效果。
\end{cnenumerate}
\par
}

\vspace{0.8em}
% 关键词：与条标题字号字体相同（小四号黑体），左对齐；3－5个，中文逗号分隔，最后一个后不打标点
\begin{flushleft}
\xiaosihao\heiti 关键词：\xiaosihao\songti 联邦学习，异构图推荐，对比学习，冷启动，模型压缩
\end{flushleft}
% --- 英文摘要：标题与章标题一致（三号加粗居中），正文与正文部分一致 ---
\newpage
\markboth{ABSTRACT}{ABSTRACT}
\vspace{\AbstractTopVSpace}
\begin{center}
    \centering
    {\sanhao\heiti ABSTRACT}
\end{center}
\vspace{\AbstractTitleBodyVSpace}

\phantomsection
\addcontentsline{toc}{section}{ABSTRACT}

% 英文摘要：顶格、无首行缩进；(1)(2)(3) 与后续英文无空格紧接；西文 Times New Roman（随 \setmainfont）
\begingroup
\rmfamily
\setlength{\parindent}{0pt}
\setlength{\parskip}{0.18\baselineskip plus 0.05\baselineskip minus 0.03\baselineskip}
{\fontsize{12pt}{18pt}\selectfont
\noindent With the rapid development of the mobile internet, recommendation systems have shifted from mere traffic distribution to deep interest mining of users. Heterogeneous information networks (HINs) have emerged as a practical technology to improve recommendation accuracy by integrating diverse heterogeneous data sources and mining hidden semantic information. Meanwhile, under the compliance requirements of the Personal Information Protection Law and GDPR, as well as the increasing ``data silos'' between major platforms, the traditional centralized large-graph training paradigm can no longer meet the demands of cross-domain data fusion and privacy compliance. Federated learning (FL), through a distributed training paradigm of ``models move while data stays'', allows participants to retain raw data locally and only exchange privacy-protected model updates (e.g., parameters or gradients under secure aggregation), thus satisfying privacy compliance requirements. However, due to the characteristics of the federated scenario, existing research on federated heterogeneous graph recommendation still faces the cold start problem, recommendation performance constrained by non-independent and identically distributed (Non-IID) data, and communication bottlenecks and deployment difficulties brought by high-dimensional heterogeneous models.

\noindent This paper addresses the above issues from three perspectives: recommendation effectiveness, communication efficiency, and engineering deployment. To mitigate performance degradation caused by cold start and Non-IID data, it proposes the federated heterogeneous graph recommendation framework FedASCL, which compensates missing semantics for zero-interaction users via attribute--structure dual-view contrastive learning on clients and suppresses client drift via global semantic prototype alignment on the server. To alleviate edge--cloud communication bottlenecks, it further proposes a semantics-aware model parameter compression strategy that trades acceptable accuracy loss for bandwidth savings at approximately $15.2\times$ communication reduction. Finally, it designs and implements a federated paper recommendation prototype for academic scenarios, integrating FedASCL and the compression strategy into an edge--cloud collaborative pipeline. Algorithm-level offline validation is conducted on public benchmarks including MovieLens-1M, Yelp, and ACM, while system-level functional testing and communication simulation verify engineering feasibility.

\noindent Specifically, the research work of this paper is as follows:

\noindent (1)To tackle the poor recommendation performance caused by cold start and Non-IID data, this paper proposes a federated heterogeneous graph recommendation framework based on federated attribute--structure contrastive learning (FedASCL). The framework employs a cross-view contrastive learning mechanism to align user attribute and graph structure semantics in the latent space and reconstructs missing topological information using attribute features; it also introduces a global semantic prototype alignment mechanism to correct client drift. Experiments show that FedASCL achieves about 3\%--8\% gains over the second-best baseline under cold start, about 8.2\% NDCG improvement at $\alpha=0.1$ over the second-best baseline, and only about 5.4\% NDCG decay relative to IID in Chapter~3.

\noindent (2)To address the problems of large parameter size and low communication efficiency in federated heterogeneous graph models, this paper proposes a semantics-aware model parameter compression strategy. This strategy designs a lightweight meta-path selector to dynamically eliminate redundant channels with low contribution to users during local training, and combines residual gradient quantization technology to compress transmitted data. Experimental results on MovieLens-1M, Yelp, and ACM show that at approximately $15.2\times$ communication reduction, Recall@20 drops by about 2.4\%, 3.5\%, and 3.2\% on the three datasets, respectively, relative to the full model, while communication overhead is reduced by approximately one order of magnitude (see Chapter~4), alleviating the communication bottleneck in federated recommendation systems.

\noindent (3)To support engineering deployment, this paper constructs a federated paper recommendation prototype with an edge-cloud layered architecture (data, algorithm engine, services, and presentation). Federated training follows synchronous rounds with secure aggregation as in Chapters~3--4; asynchronous job scheduling may be used at the service layer (e.g., task queues). A configurable local differential privacy module is integrated; ranking accuracy is validated offline in Chapter~3, while Chapter~5 focuses on integration and communication simulation.
\par
}
\endgroup

\vspace{\AbstractBodyKeywordVSpace}
% Keywords：``Keywords:'' 与关键词均为小四 Times New Roman，不加粗，左对齐
\begin{flushleft}
{\xiaosihao\rmfamily Keywords: }\xiaosihao\rmfamily Federated Learning, Heterogeneous Graph Recommendation, Contrastive Learning, Cold Start, Model Compression
\end{flushleft}
